---
title: "Thesis Fama French factors"
author: "Jesse Keränen"
date: "11/4/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

library(tidyverse)
library(lubridate)
library(ggplot2)
library(data.table)
library(tikzDevice)
library(forecast)
library(gridExtra)
```

In this file I will construct Fama French five factor model factors augmented with momentum factor. Later these factors will be used to test is machine learning portfolios provide alpha that can not be explained by established Fama French factors. factors will be calculated from exactly the same data set as is used in other parts of the study. Therefore, we start by loading the dataset constructed in Data_collection.Rmd file.
```{r}
load(file = "Data/data_w_microcap.Rdata")
```

Company characteristics calculation. We need to perform some calculations with raw variables we have downloaded from Datastream. Investment variable will be defined as annual change in total assets (WC02999). Book equity is common equity (WC03501) plus deferred taxes (WC03263). Book-to-market value will be then calculated by dividing book value by market value from last December. Profitability is defined as net income before extra items/preferred dividends (WC01551) divided by book equity. Momentum is defined as cumulated return from t-12 to t-2.
```{r}
# data <- data[Country == "Sweden"]
setorder(data, Date, -L.USD.MV)

data[, INV := (WC02999 - shift(WC02999, 12)) / shift(WC02999, 12), by = Company] # Investment
data[, BE := WC03501 + WC03263, by = Company] # Book equity
data[, BEME := BE / (DEC.MV * 1000) , by = Company] # Book to Market equity
data[, OP := WC01551 / BE, by = Company] # Profitability
data[, MOM12 := shift(USD.RI, 2, type = "lag")/shift(USD.RI, 12, type = "lag") - 1,
   by = Company] # Momentum
```

Next we want to allocate the stocks in our dataset to portfolios based on their size. We follow Fama and French (1993) and use two size portfolios. Since on average market value of Nordic stocks is much smaller than in the dataset of inital Fama French (1993) study, we follow Fama and French (2008 confirm) and consider following logic for size allocation. First for each day stocks will be sorted based on their market value in decreasing order. Then we will start assigning stocks to "Big" portfolio until combined market value of stocks in "Big portfolio reaches 90% of the whole market capitalization of the dataset in that given date. Rest of the companies are considered to be small.
```{r}
quantiles <- c(0, 0.9, 1)
labels <- c("Big", "Small")

setorder(data, Date, -L.USD.MV) # Order dataset first by date and then by market value
data[, hcol := ifelse(month >= 7, year, year-1)]
helpdt <- data[month == 7 & !is.na(L.USD.MV)]

helpdt[, aggregated_weight := cumsum(L.USD.MV)/sum(L.USD.MV), by = year]
helpdt[, temp := cut(aggregated_weight, breaks = quantiles, labels = labels)]
helpdt <- helpdt[, .(year, Company, temp, aggregated_weight)]
  
setnames(helpdt, "temp", "pf.size")
data <- merge(data, helpdt, by.x = c("Company", "hcol"), by.y = c("Company", "year"), all.x = T)

ggplot(na.omit(data[Date %in% c("1996-07-31", "2004-07-30", "2013-07-31", "2022-07-29")]),
       aes(as.factor(pf.size), L.USD.MV)) + geom_boxplot() + 
        facet_wrap(~lubridate::year(Date), scales = "free") + scale_y_log10()

ggplot(data[pf.size == "Small" & month == 7, .SD[which.max(L.USD.MV), .(L.USD.MV, pf.size)], by = Date], aes(Date, L.USD.MV)) + geom_line()
```

Next we want to allocate stock again to portfolio based on other variables and independently of the size allocation. Slightly different names will be used for portfolio formed based on different variables, but the breakpoints will be the same for each of the variables. We follow Fama and French (1993, 2015) and assign stock to "low" portfolio is its value in corresponding variable is below 30th percentile breakpoint and to "high" portfolio if the value is above 70th percentile breakpoint. All stocks between are considered neutral.

For momentum we follow slightly different process than for the other factors. First, momentum portfolio will be re-allocated monthly. Second difference is that when we calculate the long-short returns for the momentum factor we use equal weighted returns of the underlying stocks. Momentum factor is updated monthlty.
```{r}
labels <- c("Loser", "Neutral_mom", "Winner")
quantiles <- c(0.3, 0.7)
data[year > 1991, pf.mom := cut(MOM12, breaks = c(-Inf, quantile(.SD[pf.size == "Big", MOM12], quantiles, 
                                                      na.rm = T), Inf), labels = labels, na.rm = T), by = Date]
```

Let's try to plot the data to see if the distribution worked.
```{r}
ggplot(na.omit(data[Date %in% c("1996-07-31", "2004-07-30", "2013-07-31", "2021-07-30")]),
       aes(as.factor(pf.mom), MOM12)) + geom_boxplot() + facet_wrap(~lubridate::year(Date), scales = "free") + scale_y_log10()
```


Next we will repeat the process for the rest of the variables, but re-allocating only annually. Re-allocation is done each July. For each of the variables total six portfolio are formed via two fold allocation. For all variables breakpoints when defining breakpoints for certain variable for certain date only "Big" companies are considered. Then these breakpoints calculated from "Big" company data will be used to allocate all stocks to portfolios. 
```{r}
factor <- function(dt, variable, name, labels, quantiles) {
  helpdt <- dt[month==7 & year > 1991, .(temp = cut(.SD[, get(variable)], breaks = 
                  c(-Inf, quantile(.SD[pf.size =="Big", get(variable)], quantiles, na.rm = T), Inf), labels = labels), Company), by = year]
  
  setnames(helpdt, "temp", name)
  # Merge the variable allocation back from July Y to June Y+1
  dt <- merge(dt, helpdt, by.x=c("hcol", "Company"), by.y=c("year", "Company"), all.x=T)
  
  return(dt)
}

data <- factor(data, "BEME", "pf.bm", c("Low", "Neutral", "High"), quantiles)
data <- factor(data, "OP", "pf.op", c("Weak", "Neutral_op", "Robust"), quantiles)
data <- factor(data, "INV", "pf.inv", c("Conservative", "Neutral_inv", "Agressive"), quantiles)

# Form the 2x3 portfolios combining size with one additional variable
data[, Portfolio1 := paste0(pf.size,".",pf.bm)]
data[, Portfolio2 := paste0(pf.size,".",pf.op)]
data[, Portfolio3 := paste0(pf.size,".",pf.inv)]
data[, Portfolio4 := paste0(pf.size,".",pf.mom)]


#data[, L.USD.MV := ifelse(L.USD.MV > quantile(L.USD.MV, 0.999, na.rm = T), 
#                     quantile(L.USD.MV, 0.999, na.rm = T), L.MV), by = Date]

ggplot(na.omit(data[lubridate::month(Date) == 7, .(INV = mean(INV)), by = c("Date", "pf.inv")]), 
       aes(Date, INV, color = pf.inv)) + geom_line()

ggplot(na.omit(data[Date == as.POSIXct("2016-07-29", tz="UTC")]), 
       aes(L.USD.MV, BEME, color = Portfolio1)) + geom_point() + scale_y_log10() + scale_x_log10()
```

Once stocks are allocated to the portfolios we can calculate portfolio returns for each portfolio. Except for momentum, portfolio returns are value weighted returns. 
```{r}
# Function for calculating portfolio returns
portfolioreturns <- function(dt, portfolio){
  portfolio_returns <- dt[, .(USD.RET = weighted.mean(USD.RET, L.USD.MV, na.rm = T)),
                          by = c("Date", portfolio)]
  return(dcast(portfolio_returns, paste("... ~ ", portfolio), value.var = "USD.RET"))
}

portfolio_returns1 <- portfolioreturns(data, "Portfolio1")
portfolio_returns2 <- portfolioreturns(data, "Portfolio2")
portfolio_returns3 <- portfolioreturns(data, "Portfolio3")
portfolio_returns4 <- portfolioreturns(data, "Portfolio4")

# Equal weighted returns for momentum factor
#portfolio_returns4 <- data[, .(USD.RET = mean(USD.RET, na.rm = T)),
#                                    by = c("Date", "Portfolio4")]
#portfolio_returns4 <- dcast(portfolio_returns4, ... ~ Portfolio4, value.var = "USD.RET")
```

To have data in neat format all portfolio return data.tables are merged in to one big data.table.
```{r}
portfolio_returns <- merge(portfolio_returns1, portfolio_returns2, by = "Date")
portfolio_returns <- merge(portfolio_returns, portfolio_returns3, by = "Date")
portfolio_returns <- merge(portfolio_returns, portfolio_returns4, by = "Date")

portfolio_returns[is.na(portfolio_returns)] <- 0
```

Once portfolio return for each portfolio is calculated we can calculate actual long-short factor returns. Factors returns will be calculated by going short on "Low" portfolios and long on "High" portfolios. Since for each variable (excl. size) we have two "Low" and two "High" portfolio, the final "Low" return will be calculated as an average of the two "Low" portfolios and the "High" return as an average of the two "High" portfolios. For size there is total 24 portfolios. First size factor return will be calculated separately for each variable and then average of four size factor returns is calculated to obtain final size return.
```{r}
RF <- as.data.table(read_csv("Data/F-F_Research_Data_Factors.csv"))
RF[, `:=` (year = as.numeric(substr(Date, 1, 4)), month = as.numeric(substr(Date, 5, 6)), RF = RF / 100)]

# Market risk factor
market_return <- data[!is.na(USD.RET) & !is.na(L.USD.MV), .(MKT.RET = weighted.mean(USD.RET, L.USD.MV)), by = "Date"]
market_return[, `:=` (year = lubridate::year(Date), month = lubridate::month(Date))]
market_return <- merge(market_return, RF[, .(year, month, RF)], by = c("year", "month"))
RMRF_factor <- market_return[, .(Date, RMRF = MKT.RET-RF)]

# Calculate factors from portfolios
factors <- portfolio_returns[, .(Date, 
          SMBbm = (Small.High + Small.Neutral + Small.Low)/3 - (Big.High + Big.Neutral + Big.Low)/3,
          SMBop = (Small.Robust + Small.Neutral_op + Small.Weak)/3 - (Big.Robust + Big.Neutral_op +
                                                                        Big.Weak)/3,
          SMBinv = (Small.Conservative + Small.Neutral_inv + Small.Agressive)/3 - (Big.Conservative +
                                                                    Big.Neutral_inv + Big.Agressive)/3,
          SMBmom = (Small.Winner + Small.Neutral_mom + Small.Loser)/3 - (Big.Winner +
                                                                    Big.Neutral_mom + Big.Loser)/3,
          HML = (Small.High + Big.High)/2 - (Small.Low + Big.Low)/2,
          RMW = (Small.Robust + Big.Robust)/2 - (Small.Weak + Big.Weak)/2,
          CMA = (Small.Conservative + Big.Conservative)/2 - (Small.Agressive + Big.Agressive)/2,
          MOM = (Small.Winner + Big.Winner)/2 - (Small.Loser + Big.Loser)/2)
          ]

factors[, SMB := (SMBbm + SMBop + SMBinv + SMBmom)/4]

# Remove unnecessary columns
factors[, SMBop := NULL]
factors[, SMBbm := NULL]
factors[, SMBinv := NULL]
factors[, SMBmom := NULL]

# Merge market factor with other factors
factors_wide <- merge(factors, RMRF_factor, by="Date")

# Melt for plotting
factors <- melt(factors_wide, id.vars = "Date", variable.name = "Factor", value.name = "USD.RET")

# Check the factor statistics
min_max <- factors[, .(Min = min(USD.RET), Max = max(USD.RET)), by = Factor]
factor_statistics <- factors[, t.test(USD.RET, ), by = Factor]
factor_statistics[, Count := seq(.N), by = Factor][, Count := ifelse(Count == 1, "Lower Bound", "Upper Bound")]
factor_statistics <- dcast(factor_statistics, ... ~ Count, value.var = "conf.int")
factor_statistics <- merge(factor_statistics, min_max, by = "Factor")
round <- colnames(factor_statistics[, -c("Factor", "Parameter", "alternative", "method", "data.name")])
factor_statistics[, (round) := round(.SD, 4), .SDcols = round]

# Factor correlations
factor_correlations <- round(cor(factors_wide[, -c("Date")]), 4)

# Just for plotting
factors2 <- factors[lubridate::year(Date) > 1991, .(Date, cum_prod = cumprod(USD.RET + 1)), by = Factor]

tikz(file = "Latex/R_graphs/factor_performance.tex", width = 6, height = 4)
plot <- ggplot(factors2, aes(Date, cum_prod)) + geom_line(aes(color = Factor)) +
  labs(y = "Cumulative return", title = "") + scale_y_log10() +
  theme(axis.title = element_blank(), legend.position = "bottom")
print(plot)
dev.off()

plot
```

```{r}
tikz(file = "Latex/R_graphs/tes.tex", width = 6, height = 5)
plot2 <- grid.arrange(factors_wide[, ggAcf(HML)] + 
                        facet_wrap(facets = as.factor("HML")) + ggtitle(NULL) + 
                        theme(axis.title = element_blank()),
             factors_wide[, ggAcf(RMW)] + facet_wrap(facets = as.factor("RMW")) +
               ggtitle(NULL) + theme(axis.title = element_blank()),
             factors_wide[, ggAcf(CMA)] + facet_wrap(facets = as.factor("CMA")) +
               ggtitle(NULL) + theme(axis.title = element_blank()),
             factors_wide[, ggAcf(MOM)] + facet_wrap(facets = as.factor("MOM")) +
               ggtitle(NULL) + theme(axis.title = element_blank()),
             factors_wide[, ggAcf(SMB)] + facet_wrap(facets = as.factor("SMB")) +
               ggtitle(NULL) + theme(axis.title = element_blank()),
             factors_wide[, ggAcf(RMRF)] + facet_wrap(facets = as.factor("RMRF")) +
               ggtitle(NULL) + theme(axis.title = element_blank()),
             bottom = "Lag", left = "ACF")
print(plot2)
dev.off()
```


Final factor returns are again stored for later usage.
```{r}
save(factors, file = "Data/factors.Rdata")
```


```{r}
var <- "MOM"

factors3 <- copy(factors)
colnames(factors3)[3] <- "RET"
hanauer <- as.data.table(read_csv("global_factor_premia_data.csv", show_col_types = F))
hanauer <- hanauer[variable == var & market == "SWE"]
factors3[, `:=` (Date = as.Date(Date), Factor = as.character(Factor))]
factors3 <-factors3[Date >= "1992-07-01" & Factor == var]
hanauer[, `:=` (Date = as.Date(Date), Factor = as.character(variable))]
hanauer <- hanauer[Date >= "1992-07-01"]
hanauer[, return := return / 100]

hanauer <- merge(factors3, hanauer, by = c("Date", "Factor"))


hanauer[, cor(return, RET)]


hanauer <- melt(hanauer[, -c("market", "Factor", "variable")], id.vars = "Date")
hanauer[, temp := cumprod(value + 1), by = variable]

ggplot(hanauer, aes(Date, temp, color = variable)) + geom_line()
```

